---
title: "Lab 6"
author: "Max Settineri"
date: "2023-02-16"
output: 
  html_document:
    code_folding: show
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(janitor)
library(palmerpenguins)

# for cluster analysis
library(NbClust)
library(cluster)
library(factoextra)
library(dendextend)
library(ggdendro)
```

# Intro to cluster analysis - k-means, hierarchical

## Part 1: K-means clustering

```{r}
ggplot(penguins) +
  geom_point(aes(x = bill_length_mm, 
                 y = bill_depth_mm, 
                 color = species,
                 shape = sex),
             size = 3,
             alpha = 0.7) +
  scale_color_manual(values = c('orange', 'cyan4', 'darkmagenta'))

ggplot(penguins) +
  geom_point(aes(x = flipper_length_mm,
                 y = body_mass_g,
                 color = species,
                 shape = sex),
             size = 3,
             alpha = 0.7) +
  scale_color_manual(values = c('orange', 'cyan4', 'darkmagenta'))

```

### Create a complete, scaled version of the data

```{r}
penguins_complete <- penguins %>% 
  drop_na(bill_length_mm, bill_depth_mm, body_mass_g, flipper_length_mm)

penguins_scale <- penguins_complete %>% 
  select(ends_with('_mm'), body_mass_g) %>% 
  scale() # creating a matrix where all values for each column have a mean of 0 and sd of 1
```

### Estimate number of clusters

```{r}
number_est <- NbClust(penguins_scale, min.nc = 2, max.nc = 10, method = 'kmeans')

fviz_nbclust(penguins_scale, FUNcluster = kmeans, method = 'wss', k.max = 10)
```

### run some k-means

```{r}
set.seed(123)
penguins_km <- kmeans(penguins_scale,
                      centers = 3, # number of cluster centers to start with
                      iter.max = 10, # number of iterations
                      nstart = 25) # out of 25 times run, pull result with the lowest sum of squares

#penguins_km$size
#penguins_km$cluster

penguins_cl <- penguins_complete %>% 
  mutate(cluster_no = factor(penguins_km$cluster))
```

```{r}
ggplot(penguins_cl) +
  geom_point(aes(x = flipper_length_mm,
                   y = body_mass_g, 
                 color = cluster_no,
                 shape = species))

ggplot(penguins_cl) +
  geom_point(aes(x = bill_length_mm,
                   y = bill_depth_mm, 
                 color = cluster_no,
                 shape = species)) +
  scale_color_viridis_d()

penguins_cl %>% 
  select(species, cluster_no) %>% 
  table() # shows us how well clusters are grouping species
```

## Hierarchical clustering

### Start with complete linkage

```{r}
# create distance matrix
peng_dist <- dist(penguins_scale, method = 'euclidean')

# hierarchical clustering (complete linkage)
peng_hc_complete <- hclust(peng_dist, method = 'complete')
# also single, average, ward.D 

# plot dendrogram
plot(peng_hc_complete, cex = .6, hang = -1)

# cut tree into thee clusters
peng_cut_hc <- cutree(peng_hc_complete, 3)
table(peng_cut_hc, penguins_complete$species)
```

## World bank data: read in and simplify

```{r}
wb_env <- read_csv(here::here('data', 'wb_env.csv'))

wb_ghg_20 <- wb_env %>% 
  slice_max(n = 20, ghg)

summary(wb_ghg_20)

wb_scaled <- wb_ghg_20 %>% 
  select(3:7) %>% 
  scale()

summary(wb_scaled)

rownames(wb_scaled) <- wb_ghg_20$name 
```

### Find the Euclidean distances

```{r}
euc_distance <- dist(wb_scaled, method = 'euclidean')
euc_distance
```

### Perform hierarchical clustering with complete linkage

```{r}
hc_complete <- hclust(euc_distance, method = 'complete')
plot(hc_complete, cex = .6, hang = -.1)
```

### Perform hierarchical clustering by single linkage

```{r}
hc_single <- hclust(euc_distance, method = 'single')
plot(hc_single, cex = .6, hang = -.1)
```

### Make a tanglegram

```{r}
dend_complete <- as.dendrogram(hc_complete)
dend_single <- as.dendrogram(hc_single)

tanglegram(dend_complete, dend_single)

entanglement(dend_complete, dend_single)

untangle(dend_complete, dend_single, method = 'step1side') %>% 
  entanglement()

untangle(dend_complete, dend_single, method = 'step1side') %>%
  tanglegram(common_subtrees_color_branches = TRUE)
```

### Let's make a dendrogram in ggplot

```{r}
ggdendrogram(hc_complete, rotate = TRUE) +
  theme_minimal() +
  labs(x = 'Country')
```

